Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-09 CAs_operads.tex
Relecture partie 3 (avant partie 3.3)

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
    Let $\gamma$ be a positive integer and let $\Tfr$ and $\Tfr'$ be two
    binary trees. If $\Tfr \CongrCAs{\gamma} \Tfr'$, then
    \begin{equation} \label{equ:left_rank_and_CongrCAs}
        \LRank(\Tfr) \pmod{\bar{\gamma}}
        \enspace = \enspace
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}}.
    \end{equation}
\end{Lemma}
\begin{proof}
    Consider here the rewrite rule $\Rew$ on $\Mag$ satisfying
    $\LComb{\gamma} \Rew \RComb{\gamma}$. Let us show that the rewrite
    relation $\RewContext$ induced by $\Rew$ is such that
    $\Tfr \RewContext \Tfr'$ implies~\eqref{equ:left_rank_and_CongrCAs}.
    Any binary tree $\Tfr$ decomposes as
    \begin{math}
        \Tfr =
        \LComb{\LRank\left(\Tfr\right)} \circ
        \left[\Leaf, \Tfr_1, \dots,
        \Tfr_{\LRank\left(\Tfr\right)} \right]
    \end{math}
    where the $\Tfr_i$ are binary trees. Now, if
    $\Tfr \RewContext \Tfr'$, then one among the following two cases
    occurs.
    \begin{enumerate}[label={(\it\roman*)}]
        \item The rewrite step is applied into one of the trees
        $\Tfr_i$, that is there exists $\Tfr_i'$ such that
        \begin{math}
            \Tfr' = \LComb{\LRank\left(\Tfr\right)}
            \circ \left[\Leaf, \Tfr_1, \dots, \Tfr_i', \dots,
                \Tfr_{\LRank\left(\Tfr\right)}
            \right]
        \end{math}
        so that
        \begin{math}
            \LRank\left(\Tfr'\right) = \LRank\left(\Tfr\right).
        \end{math}
        \item The rewrite step is applied into the left branch
          beginning
            at the root of $\Tfr$, that is there exists $i$ such that
            \begin{math}
                \Tfr' = \LComb{\LRank\left(\Tfr\right)-\bar{\gamma}}
                \circ \left[\Leaf, \Tfr_1, \dots, \RComb{\bar{\gamma}}
                \circ
                \left[\Tfr_{i},\dots, \Tfr_{i+\bar{\gamma}}\right],
                \dots, \Tfr_{\LRank\left(\Tfr\right)} \right]
            \end{math}
            so that
            \begin{math}
                \LRank\left(\Tfr'\right) =
                \LRank\left(\Tfr\right) - \bar{\gamma} =
                \LRank\left(\Tfr\right) \pmod{\bar{\gamma}}
            \end{math}.
    \end{enumerate}
    This implies~\eqref{equ:left_rank_and_CongrCAs}. Finally, since
    $\CongrCAs{\gamma}$ is the reflexive, symmetric, and transitive
    closure of~$\RewContext$, the statement of the lemma follows.
\end{proof}
\medbreak

\begin{Proposition} \label{prop:division_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers. Then, there
    exists a morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ if and
    only if $\bar{\gamma} \mid \bar{\gamma'}$.
\end{Proposition}
\begin{proof}
    From Lemma~\ref{lem:morphism_CAs}, it is enough to show that
    $\LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}$ if and only if
    $\bar{\gamma} \mid \bar{\gamma'}$. If
    $\LComb{\gamma'}\CongrCAs{\gamma}\RComb{\gamma'}$, $\gamma$ is
    smaller than $\gamma'$ (this is a consequence of the existence of a
    surjective morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ and
    Lemma~\ref{lem:first_dimensions_CAs}). From
    Lemma~\ref{lem:left_rank_and_CongrCAs}, we deduce that
    \begin{math}
        \LRank\left(\LComb{\gamma'}\right)
        - \LRank\left(\RComb{\gamma'}\right)
        = \bar{\gamma'}
    \end{math}
    is divisible by $\bar{\gamma}$, which shows the direct implication.
    \smallbreak

    Conversely, if
    $\bar{\gamma} \mid \bar{\gamma'}$, the rewrite
    rule $\LComb{\gamma} \Rew \RComb{\gamma}$ induces the
    sequence

@@

    of rewriting steps, where squared regions denotes left or right
    comb trees of degree $\gamma - 1$. Hence, we have
    $\LComb{\gamma'}\CongrCAs{\gamma}\RComb{\gamma'}$.

@@

The lattice $\left(\N, \mid, \gcd, \lcm\right)$ admits $1$ as minimum
element and $0$ as maximum element: any nonnegative integer is
divisible by $1$ and divides $0$. The corresponding property for
$\LatCAs$ is that it admits $\As=\CAs{2}$ as minimum element and $\Mag=
\CAs{1}$ as maximum element: any comb associative operad projects onto
$\As$ and is a quotient of $\Mag$.
\medbreak

@@

Note that $\LatKCAs$ does not embed as a sublattice of $\LatQMag$, that
is $\iota$ is not a lattice morphism. Consider for instance $\gamma = 3$
and $\gamma' = 4$, so that $\KCAs{\gamma} \InfCAs \KCAs{\gamma'}$ is
equal to $\KCAs{2} = \K\As$, whereas
$\KCAs{\gamma} \InfQMag \KCAs{\gamma'}$ is the quotient of $\KMag$ by the
two relations $\LComb{3} - \RComb{3}$ and $\LComb{4} - \RComb{4}$.
\medbreak

>>>>

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
    Let $\gamma \geq 2$ be an integer and let $\Tfr$ and $\Tfr'$ be two
    binary trees. If $\Tfr \CongrCAs{\gamma} \Tfr'$, then
    \begin{equation} \label{equ:left_rank_and_CongrCAs}
        \LRank(\Tfr) \pmod{\bar{\gamma}}
        \enspace = \enspace
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}}.
    \end{equation}
\end{Lemma}
\begin{proof}
    Consider here the rewrite rule $\Rew$ on $\Mag$ satisfying
    $\LComb{\gamma} \Rew \RComb{\gamma}$. Let us show that the rewrite
    relation $\RewContext$ induced by $\Rew$ is such that
    $\Tfr \RewContext \Tfr'$ implies~\eqref{equ:left_rank_and_CongrCAs}.
    Any binary tree $\Tfr$ decomposes as
    \begin{math}
        \Tfr =
        \LComb{\LRank\left(\Tfr\right)} \circ
        \left[\Leaf, \Tfr_1, \dots,
        \Tfr_{\LRank\left(\Tfr\right)} \right]
    \end{math}
    where the $\Tfr_i$ are binary trees. Now, if
    $\Tfr \RewContext \Tfr'$, then one among the following two cases
    occurs.
    \begin{enumerate}[label={(\it\roman*)}]
        \item The rewrite step is applied into one of the trees
        $\Tfr_i$, that is there exists $\Tfr_i'$ such that
        \begin{math}
            \Tfr' = \LComb{\LRank\left(\Tfr\right)}
            \circ \left[\Leaf, \Tfr_1, \dots, \Tfr_i', \dots,
                \Tfr_{\LRank\left(\Tfr\right)}
            \right]
        \end{math}
        so that
        \begin{math}
            \LRank\left(\Tfr'\right) = \LRank\left(\Tfr\right).
        \end{math}
        \item The rewrite step is applied into the left branch
          beginning
            at the root of $\Tfr$, that is there exists $i$ such that
            \begin{math}
                \Tfr' = \LComb{\LRank\left(\Tfr\right)-\bar{\gamma}}
                \circ \left[\Leaf, \Tfr_1, \dots, \RComb{\bar{\gamma}}
                \circ
                \left[\Tfr_{i},\dots, \Tfr_{i+\bar{\gamma}}\right],
                \dots, \Tfr_{\LRank\left(\Tfr\right)} \right]
            \end{math}
            so that
            \begin{math}
                \LRank\left(\Tfr'\right) =
                \LRank\left(\Tfr\right) - \bar{\gamma} =
                \LRank\left(\Tfr\right) \pmod{\bar{\gamma}}
            \end{math}.
    \end{enumerate}
    This implies~\eqref{equ:left_rank_and_CongrCAs}. Finally, since
    $\CongrCAs{\gamma}$ is the reflexive, symmetric, and transitive
    closure of~$\RewContext$, the statement of the lemma follows.
\end{proof}
\medbreak

\begin{Proposition} \label{prop:division_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers such that
    $\gamma \geq 2$. Then, there
    exists a morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ if and
    only if $\bar{\gamma} \mid \bar{\gamma'}$.
\end{Proposition}
\begin{proof}
    From Lemma~\ref{lem:morphism_CAs}, it is enough to show that
    $\LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}$ if and only if
    $\bar{\gamma} \mid \bar{\gamma'}$. If
    \begin{math}
        \LComb{\gamma'}\CongrCAs{\gamma}\RComb{\gamma'},
    \end{math}
    as a consequence of the existence of a surjective morphism $\varphi$
    from $\CAs{\gamma'}$ to $\CAs{\gamma}$ and
    Lemmas~\ref{lem:first_dimensions_CAs}
    and~\ref{lem:surjective_morphisms_CAs}, one has $\gamma' = 1$ or
    $\gamma \leq \gamma'$. Since
    \begin{equation}
        \LRank\left(\LComb{\gamma'}\right)
        - \LRank\left(\RComb{\gamma'}\right)
        = \gamma' - 1
        = \bar{\gamma'},
    \end{equation}
    by using Lemma~\ref{lem:left_rank_and_CongrCAs}, we deduce that
    $\bar{\gamma'}$ is divisible by $\bar{\gamma}$, which shows the
    direct implication.
    \smallbreak

    Conversely, if $\bar{\gamma} \mid \bar{\gamma'}$, the rewrite rule
    $\Rew$ on $\Mag$ satisfying $\LComb{\gamma} \Rew \RComb{\gamma}$
    induces the sequence

@@

    of rewrite steps, where squared regions denotes left or right
    comb trees of degree $\gamma - 1$. Hence, since
    $\CongrCAs{\gamma}$ is the reflexive, symmetric, and transitive
    closure of $\RewContext$, we
    have~$\LComb{\gamma'}\CongrCAs{\gamma}\RComb{\gamma'}$.

@@

The lattice $\left(\N, \mid, \gcd, \lcm\right)$ admits $1$ as minimum
element and $0$ as maximum element since any nonnegative integer is
divisible by $1$ and divides $0$. These properties translate as follows
for $\LatCAs$: the minimum element is $\As=\CAs{2}$ and the maximum
element is $\Mag=\CAs{1}$. Algebraically, this says that any comb
associative operad projects onto $\As$ and is a quotient of~$\Mag$.
\medbreak

@@

Note that $\LatKCAs$ does not embed as a sublattice of $\LatQMag$, that
is $\iota$ is not a lattice morphism. Consider for instance $\gamma = 3$
and $\gamma' = 4$, so that
\begin{equation}
    \KCAs{3} \InfCAs \KCAs{4} = \KCAs{2} = \K\Angle{\As},
\end{equation}
whereas
\begin{equation}
    \KCAs{3} \InfQMag \KCAs{4} =
    \KMag/_I
\end{equation}
where $I$ is the ideal of $\KMag$ generated by
$\LComb{3} - \RComb{3}$ and $\LComb{4} - \RComb{4}$.
\medbreak

### 2018-09-09 Operads_rewrite.tex
Relecture partie 3 (avant partie 3.3)

When $\Rew$ is a convergent orientation of $\Congr$, the set
$\NormalForms_{\RewContext}$ of all normal forms of $\RewContext$ is
called a \Def{Poincaré-Birkhoff-Witt basis} (or a \Def{PBW basis} for
short) of the quotient operad $\Oca$ (see~\cite{Hof10,DK10}). This forms
a one-to-one correspondence between the sets
$\NormalForms_{\RewContext}(n)$ and $\Oca(n)$, $n \geq 1$. In other
words, a PBW basis offers a way to assign with each $\Congr$-equivalence
class $[\Tfr]_{\Congr}$ a representative
\begin{math}
    \Tfr' \in [\Tfr]_{\Congr} \cap \NormalForms_{\RewContext}.
\end{math}
A \Def{combinatorial realization} of an operad $\Oca$ of the
form~\eqref{equ:quotient_operad_mag} is an operad $\Cca$ isomorphic to
$\Oca$ which admits an explicit description of its elements and an
explicit description of its partial composition maps. The knowledge of a
PBW basis $\Cca := \NormalForms_{\RewContext}$ of $\Oca$ provides a
combinatorial realization $\Cca$ of $\Oca$. Indeed, the partial
composition $\Tfr' \circ_i \Sfr'$ of two binary trees $\Tfr'$ and
$\Sfr'$ of $\Cca$ is the tree obtained by grafting the root of $\Sfr'$
onto the $i$th leaf of $\Tfr'$ and by rewriting by $\RewContext$ this
tree as much as possible in order to obtain a normal form. This process
is well-defined since, by hypothesis, $\RewContext$ is convergent.
\medbreak

>>>>

When $\Rew$ is a convergent orientation of $\Congr$, the set
$\NormalForms_{\RewContext}$ of all normal forms of $\RewContext$ is
called a \Def{Poincaré-Birkhoff-Witt basis}~\cite{Hof10,DK10} (or a
\Def{PBW basis} for short) of the quotient operad $\Oca$. This forms a
one-to-one correspondence between the sets
$\NormalForms_{\RewContext}(n)$ and $\Oca(n)$, $n \geq 1$. In other
words, a PBW basis offers a way to assign with each $\Congr$-equivalence
class $[\Tfr]_{\Congr}$ a representative
\begin{math}
    \Tfr' \in [\Tfr]_{\Congr} \cap \NormalForms_{\RewContext}.
\end{math}
A \Def{combinatorial realization} of an operad $\Oca$ of the
form~\eqref{equ:quotient_operad_mag} is an operad $\Cca$ isomorphic to
$\Oca$ which admits an explicit description of its elements and an
explicit description of its partial composition maps. The knowledge of a
PBW basis $\Cca := \NormalForms_{\RewContext}$ of $\Oca$ provides a
combinatorial realization $\Cca$ of $\Oca$. Indeed, the partial
composition $\Tfr' \circ_i \Sfr'$ of two binary trees $\Tfr'$ and
$\Sfr'$ of $\Cca$ is the tree obtained by grafting the root of $\Sfr'$
onto the $i$th leaf of $\Tfr'$ and by rewriting by $\RewContext$ this
tree as much as possible in order to obtain a normal form. This process
is well-defined since, by hypothesis, $\RewContext$ is convergent.
\medbreak
